Supplementary information - The Functional Therapeutic Chemical Classification System

### 2013-08-18 document.tex
references added
Signed-off-by: loopasam <samuel.croset@gmail.com>

\section{Source code}
The code behind the creation of the resource is entirely open and available 
at {{https://github.com/loopasam/ftc}}. The web application built on the top of the 
FTC can be find at {{https://www.ebi.ac.uk/chembl/ftc}} and the documentation can be 
accessed at {{https://github.com/loopasam/ftc/wiki}}. The reader should be familiar with 
description logics and the Web Ontology Language (OWL) to fully understand the construction of the 
knowledge base. An introduction to description logics from the perspective of the biomedical scientist is 
available on the wiki at {{https://github.com/loopasam/ftc/wiki/Description-Logics}}. The FTC implementation 
relies mostly on Brain (cite) and the web application builds on the top of the Play! framework (cite). 
Classification tasks use the ELK reasoner (cite). The computer hosting the web application has 8 Gb of memory 
with 4 processors, this architecture allows fast parallel reasoning, thanks to ELK’s design. More functionalities 
will be added to the web application following user requirements (lean implementation).

\section{FTC categories creation}
The mode of action categories present in the FTC are defined based on the terms coming from the 
Gene Ontology (GO). Both the molecular function and biological process branches are used for this purpose, yet handled differently.

@@

\subsection{DrugBank}
The DrugBank database is a unique bioinformatics and cheminformatics resource 
that combines detailed drug data with comprehensive drug target information (cite). The approved 
drugs (small molecules and biotherapeutics) acting on proteins are extracted from the database and 
imported in the FTC knowledge base. In order to be selected, a compound must firstly be approved and 
secondly have a pharmacological action on at least one human protein target present in Uniprot (cite). 
The protein targets all have at least one manually asserted GO annotation for a biological process or a 
molecular function. DrugBank links compounds to targets via actions. The DrugBank actions are somehow 
structured and consistent: Concepts such as inhibitor or agonist are reused throughout the database for 
example, yet they are not strictly formalised as a controlled vocabulary. These actions are manually 
standardised to the core properties of the FTC according to their meaning: For instance the action 
antagonist is mapped to the FTC negatively-perturbs.
Compounds coming from DrugBank are represented as OWL classes and asserted 
as subclasses of DrugBank compound (FTC\_C2). Protein targets are described as classes 
too and subclasses of the core class protein. Each DrugBank compound is then connected to 
its target via the following axiom pattern: drug SubClassOf perturbs some protein. 
E.g. Ximelagatran SubClassOf negatively-perturbs some Prothrombin.

\subsection{Gene Ontology annotations}
The GO annotation program aims to provide high-quality GO annotations to 
proteins in UniProt (cite). In the context of the FTC, such annotations are used 
to create axioms linking protein targets to molecular functions and biological processes. 
Each protein annotated with a function creates an axiom such 
as protein SubClassOf has-function some molecular function. 
Each protein annotated to a biological process creates an axiom such 
as protein SubClassOf involved-in some biological process. 
E.g. Prothrombin SubClassOf involved-in some positive regulation of blood coagulation. 
Each protein can be involved in multiple processes and capable of realizing multiple functions.

\section{Knowledge base classification}Supplementary Material
The knowledge base is fully built at this step and contains core classes, mode 
of actions descriptions alongside the actions of approved DrugBank compounds on 
protein targets in Uniprot. The proteins are linked to their molecular functions and involvement in 
biological processes via the GO annotations. The logical specifications of the FTC are there to glue 
the different data together and to explicitly express the logical links between resources. The FTC 
knowledge base follows an OWL2 EL profile, which enable the use of fast and parallelised reasoners such 
as ELK (cite). During the classification process, the reasoner checks whether the mode of action equivalent 
definitions are satisfied or not and assigns drugs inside the corresponding FTC categories. The tree structure 
of FTC appears also at this step from the logical definitions.

\section{Evaluation methodology}
As the classification of therapeutic agents is done in an automated way, it 
is important to evaluate the results generated against a known resource which will be 
considered as gold standard. The assessment of the FTC is done against another similar 
classification, the Anatomical Therapeutic Chemical Classification System (ATC) (cite). 
The ATC has been developed to serve as a tool for drug utilization research in order to 
improve quality of drug use. In this resource, the information is manually curated, and drugs 
are assigned to categories based of their legally approved indications. The goal of the ATC differs 
from the one of the FTC, yet the two resources are sharing some very similar concepts, which can be 
used for the evaluation. Categories of both classifications contain approved drugs with a DrugBank 
identifier, meaning that some of the drugs indexed in the FTC are also present in the ATC. From that, 
it is possible to define some evaluation points, which will help to assess the automated classification process.

@@

\section{Semantic similarity}
The semantic similarity measure performed over the FTC is a derivative of the Jaccard index (cite). 
It is probably best understood as an example: If we consider two classes A and B, the semantic similarity 
between these classes corresponds to the number of OWL superclasses (direct and indirect, obtained with a reasoner) that 
are shared by A and B (intersection) divided by the number of superclasses of A or B (union). The index ranges 
from 0 (totally different) to 1 (identical).

@@

%\bibliographystyle{natbib}
%\bibliographystyle{achemnat}
%\bibliographystyle{plainnat}
%\bibliographystyle{abbrv}
%\bibliographystyle{bioinformatics}
%
%\bibliographystyle{plain}
%
%\bibliography{Document}

\begin{thebibliography}{}
\bibitem[Bofelli {\it et~al}., 2000]{Boffelli03} Bofelli,F., Name2, Name3 (2003) Article title, {\it Journal Name}, {\bf 199}, 133-154.

\bibitem[Bag {\it et~al}., 2001]{Bag01} Bag,M., Name2, Name3 (2001) Article title, {\it Journal Name}, {\bf 99}, 33-54.

\bibitem[Yoo \textit{et~al}., 2003]{Yoo03}
Yoo,M.S. \textit{et~al}. (2003) Oxidative stress regulated genes
in nigral dopaminergic neurnol cell: correlation with the known
pathology in Parkinson's disease. \textit{Brain Res. Mol. Brain
Res.}, \textbf{110}(Suppl. 1), 76--84.

\bibitem[Lehmann, 1986]{Leh86}
Lehmann,E.L. (1986) Chapter title. \textit{Book Title}. Vol.~1, 2nd edn. Springer-Verlag, New York.

\bibitem[Crenshaw and Jones, 2003]{Cre03}
Crenshaw, B.,III, and Jones, W.B.,Jr (2003) The future of clinical
cancer management: one tumor, one chip. \textit{Bioinformatics},
doi:10.1093/bioinformatics/btn000.

\bibitem[Auhtor \textit{et~al}. (2000)]{Aut00}
Auhtor,A.B. \textit{et~al}. (2000) Chapter title. In Smith, A.C.
(ed.), \textit{Book Title}, 2nd edn. Publisher, Location, Vol. 1, pp.
???--???.

\bibitem[Bardet, 1920]{Bar20}
Bardet, G. (1920) Sur un syndrome d'obesite infantile avec
polydactylie et retinite pigmentaire (contribution a l'etude des
formes cliniques de l'obesite hypophysaire). PhD Thesis, name of
institution, Paris, France.

\end{thebibliography}
\end{document}

>>>>

\section{Source code}
The code behind the creation of the resource is entirely open and available 
at {{https://github.com/loopasam/ftc}}. The web application built on the top of the 
FTC can be find at {{https://www.ebi.ac.uk/chembl/ftc}} and the documentation can be 
accessed at {{https://github.com/loopasam/ftc/wiki}}. The reader should be familiar with 
description logics and the Web Ontology Language (OWL) to fully understand the construction of the 
knowledge base. An introduction to description logics from the perspective of the biomedical scientist is 
available on the wiki at {{https://github.com/loopasam/ftc/wiki/Description-Logics}}. The FTC implementation 
relies mostly on Brain \citep{Croset2013} and the web application builds on the top of the Play! framework. 
Classification tasks use the ELK reasoner \citep{Kazakov2011}. The computer hosting the web application has 8 Gb of memory 
with 4 processors, this architecture allows fast parallel reasoning, thanks to ELK’s design. More functionalities 
will be added to the web application following user requirements (lean implementation).

\section{FTC categories creation}
The mode of action categories present in the FTC are defined based on the terms coming from the 
Gene Ontology (GO) \citep{Ashburner2000}. Both the molecular function and biological process branches are used for 
this purpose, yet handled differently.

@@

\subsection{DrugBank}
The DrugBank database is a unique bioinformatics and cheminformatics resource 
that combines detailed drug data with comprehensive drug target information \citep{Knox2011}. The approved 
drugs (small molecules and biotherapeutics) acting on proteins are extracted from the database and 
imported in the FTC knowledge base. In order to be selected, a compound must firstly be approved and 
secondly have a pharmacological action on at least one human protein target present in Uniprot \citep{TheUniprotConsortium2013}. 
The protein targets all have at least one manually asserted GO annotation \citep{Dimmer2012} for a biological process or a 
molecular function. DrugBank links compounds to targets via actions. The DrugBank actions are somehow 
structured and consistent: Concepts such as inhibitor or agonist are reused throughout the database for 
example, yet they are not strictly formalised as a controlled vocabulary. These actions are manually 
standardised to the core properties of the FTC according to their meaning: For instance the action 
antagonist is mapped to the FTC negatively-perturbs.
Compounds coming from DrugBank are represented as OWL classes and asserted 
as subclasses of DrugBank compound (FTC\_C2). Protein targets are described as classes 
too and subclasses of the core class protein. Each DrugBank compound is then connected to 
its target via the following axiom pattern: \emph{drug SubClassOf perturbs some protein}. 
E.g. \emph{Ximelagatran SubClassOf negatively-perturbs some Prothrombin}.

\subsection{Gene Ontology annotations}
The GO annotation program aims to provide high-quality GO annotations to 
proteins in UniProt. In the context of the FTC, such annotations are used 
to create axioms linking protein targets to molecular functions and biological processes. 
Each protein annotated with a function creates an axiom such 
as protein SubClassOf has-function some molecular function. 
Each protein annotated to a biological process creates an axiom such 
as protein SubClassOf involved-in some biological process. 
E.g. Prothrombin SubClassOf involved-in some positive regulation of blood coagulation. 
Each protein can be involved in multiple processes and capable of realizing multiple functions.

\section{Knowledge base classification}
The knowledge base is fully built at this step and contains core classes, mode 
of actions descriptions alongside the actions of approved DrugBank compounds on 
protein targets in Uniprot. The proteins are linked to their molecular functions and involvement in 
biological processes via the GO annotations. The logical specifications of the FTC are there to glue 
the different data together and to explicitly express the logical links between resources. The FTC 
knowledge base follows an OWL2 EL profile, which enable the use of fast and parallelised reasoners such 
as ELK (cite). During the classification process, the reasoner checks whether the mode of action equivalent 
definitions are satisfied or not and assigns drugs inside the corresponding FTC categories. The tree structure 
of FTC appears also at this step from the logical definitions.

\section{Evaluation methodology}
As the classification of therapeutic agents is done in an automated way, it 
is important to evaluate the results generated against a known resource which will be 
considered as gold standard. The assessment of the FTC is done against another similar 
classification, the Anatomical Therapeutic Chemical Classification System (ATC) \citep{world2000anatomical}). 
The ATC has been developed to serve as a tool for drug utilization research in order to 
improve quality of drug use. In this resource, the information is manually curated, and drugs 
are assigned to categories based of their legally approved indications. The goal of the ATC differs 
from the one of the FTC, yet the two resources are sharing some very similar concepts, which can be 
used for the evaluation. Categories of both classifications contain approved drugs with a DrugBank 
identifier, meaning that some of the drugs indexed in the FTC are also present in the ATC. From that, 
it is possible to define some evaluation points, which will help to assess the automated classification process.

@@

\section{Semantic similarity}
The semantic similarity measure performed over the FTC is a derivative of the Jaccard index \citep{jaccard1912distribution} \citep{rogers1960computer}. 
It is probably best understood as an example: If we consider two classes A and B, the semantic similarity 
between these classes corresponds to the number of OWL superclasses (direct and indirect, obtained with a reasoner) that 
are shared by A and B (intersection) divided by the number of superclasses of A or B (union). The index ranges 
from 0 (totally different) to 1 (identical).

@@

\begin{figure*}[!tpb]%figureS4
\centerline{\includegraphics{figS4.png}}
\caption{Caption to be written.}\label{fig:S04}
\end{figure*}

\begin{figure*}[!tpb]%figureS5
\centerline{\includegraphics{figS5.png}}
\caption{Caption to be written.}\label{fig:S05}
\end{figure*}

\bibliographystyle{natbib}
\bibliographystyle{achemnat}
\bibliographystyle{plainnat}
\bibliographystyle{abbrv}
\bibliographystyle{bioinformatics}

\bibliographystyle{plain}

\bibliography{document}

\end{document}
